Exploration of TSA Claims Data
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

setwd("~/Udacity/eda/Project")
library(ggplot2)
library(lubridate)
library(knitr)
library(gridExtra)
library(dplyr)
library(ggmap)
library(GGally)
library(scales)
library(memisc)
library(waffle)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, 
                      message=FALSE)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
tsa <- read.csv('tsa_claims.csv')
long_lat <- read.csv('airports.csv')
```

---
title: "Exporation of TSA Claims"
author: Quentin Ikeno
date: [September] [23], [2017]
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

#Introduction

The Transportation Security Administration (TSA) is an agency of the U.S. 
Department of Homeland Security, responsible for the public saftey of travelers
in the United States.

Here we will be exploring claims data against the TSA from 2002 to 2015.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
str(tsa)
#summary(tsa)
```

This dataframe contains 204,267 obvservations of 13 variables.

```{r, echo=FALSE}
tsa$date.received.format <- as.Date(tsa$Date.Received, format = "%d-%b-%y")
tsa$date.received.year <- as.numeric(format(tsa$date.received.format, format = "%Y"))
tsa$date.received.month <- as.numeric(format(tsa$date.received.format, format = "%m"))
tsa$date.received.day <- as.numeric(format(tsa$date.received.format, format = "%d"))

#convert incident date to a single format
tsa$incident.date.format <- ymd_hms(tsa$Incident.Date)
a <- as.Date(tsa$Incident.Date, format=("%m/%d/%Y %H:%M"))
b <- as.Date(tsa$Incident.Date, format=("%d-%b-%y"))
b[is.na(b)] <- a[!is.na(a)]
tsa$incident.date.format <- b


#fill in missing disposition data
for (i in 1:length(tsa$Disposition)) {
  if (tsa$Disposition[i] == '') {
    if (tsa$Status[i] == 'Deny' | tsa$Status[i] == 'Denied') {
      tsa$Disposition[i] <- 'Deny'
    }
    else if (tsa$Status[i] == 'Settle' | tsa$Status[i] == 'Settled') {
      tsa$Disposition[i] <- 'Settle'
    }
    else if (tsa$Status[i] == 'Approve in Full') {
      tsa$Disposition[i] <- 'Approve in Full'
    }
    else{}
  }
}

wrong.years <- c(96945, 96944, 96939, 96938, 96940, 96941, 96942, 96937, 	
96946, 96967, 96963, 96965, 96964, 96947, 96943)

#fix dates that are wrong
for (i in 1:length(wrong.years)) {
  if((tsa$date.received.year[wrong.years[i]] < 2002) | (tsa$date.received.year[wrong.years[i]] > 2015)) {
      tsa$incident.date.format[wrong.years[i]] <- as.Date(tsa$Incident.Date[i], format = c("%m/%d/%Y %H:%M"))
      tsa$date.received.year[wrong.years[i]] <- as.numeric(format(tsa$incident.date.format[wrong.years[i]], format = "%Y"))
      year(tsa$date.received.format[wrong.years[i]]) <- tsa$date.received.year[wrong.years[i]]
    } else {
    }
}

tsa$incident.date.year <- as.numeric(format(tsa$incident.date.format, format = "%Y"))
tsa$incident.date.month <- as.numeric(format(tsa$incident.date.format, format = "%m"))
tsa$incident.date.day <- as.numeric(format(tsa$incident.date.format, format = "%d"))

#leave only numbers in the claim amount and close amount columns
tsa$Claim.Amount <-  as.numeric(gsub("[^0-9\\.]", "", tsa$Claim.Amount))
tsa$Close.Amount <- as.numeric(gsub("[^0-9\\.]", "", tsa$Close.Amount))

sum(is.na(tsa$Close.Amount))
```

There some columns in the dataframe with missing data.  For example, there are 
68,952 missing values for close amount.  I made a quick look through the data 
and it appears that the TSA stopped recording the close amount in the claims 
data, starting from 2010.  

```{r, echo=FALSE, warning=FALSE}
qplot(data = subset(tsa, !is.na(date.received.year)),
      date.received.year,
      binwidth = 1,
      geom = 'freqpoly',
      color = I('#5760AB')) +
  scale_x_continuous(limits = c(2002, 2015),
                     breaks = seq(2002, 2015, 1)) +
  ggtitle('Claims By Year')
```

Two things about this graph caught my eye.  First, there is a relatively low 
amount of claims for 2002, with a large jump in claims after 2002.  What could 
have possibly caused this?  

One explanation could be that in 2002 claims were lower because the TSA did not 
handle the screening of passengers and baggage at every airport in America 
(screenings at airports before 2002 were handled by private security firms).  
However, that changed when the TSA was required to screen all passengers at U.S. commercial aiports by November 19, 2002 and to screen all checked baggage by 
December 31, 2002.  Subsequently, claims went up in 2003 after the policy 
changes took effect.

Another possibility is that claims were low because there fewer people flying
since air travel in the United States was down in the aftermath of September
11, 2001.

The other thing I noticed was that the number of claims appears to decrease 
after 2004 and level off around 2010.

```{r, echo=FALSE}
qplot(data = subset(tsa, !is.na(date.received.year)),
      date.received.month,
      binwidth = 1,
      color = I('black'),
      fill = I('#5760AB')) +
  scale_x_continuous(limits = c(0, 13),
                     breaks = seq(1, 12, 1)) +
  ggtitle('Claims By Month') +
  facet_wrap(~ date.received.year)
```

From this plot we can see that the number of claims started increasing around 
late 2002, peaking in 2003, then start to decrease around 2004.  There doesn't
appear to any particular month or travel season when claims increase.

```{r, echo=FALSE}
head(sort(table(tsa$Airport.Name), decreasing = T))
```

Here is a table of the top five airports where the most claims were made.  The
results aren't too surprising considering these are also some of the busiest 
airports in the U.S.  According to ABC news, each of these airports are in the 
top 20 of airports where TSA employees were fired for theft.

Below is a map of the U.S. where TSA claims were filed, where the bubble size
represents the number of claims.

```{r echo=FALSE, warning=FALSE,message=FALSE}
tsa2 <- inner_join(tsa, long_lat, by = c('Airport.Code' = 'airport.code'))
tsa.by.port <- tsa2 %>%
  group_by(Airport.Code, latitude, longitude) %>%
  summarize(n = n()) %>%
  arrange(desc(n))

#head(tsa.by.port)

map <- get_map(location = 'united states',
             zoom = 4,
             maptype = 'terrain',
             source = 'google',
             color = 'color')

ggmap(map) +
  geom_point(aes(x = longitude, y = latitude, color = I('#3300FF')),
             data = tsa.by.port,
             alpha = 0.25,
             size = tsa.by.port$n / 500) 
```


```{r, echo=FALSE}
head(sort(table(tsa$Item), decreasing = T))
```

Here are the five most common items claims were filed for.  At first I did not 
expect digital cameras to be on this list but then I remembered this data goes
back to 2002.

```{r, echo=FALSE}
qplot(data = subset(tsa, !is.na(Claim.Type) & Claim.Type != '' & Claim.Type != '-'),
      Claim.Type,
      color = I('black'),
      fill = I('#F79420')) +
  ggtitle('Claim Types') +
  coord_flip()

sort(table(tsa$Claim.Type), decreasing = TRUE)
```

I was not very surprised that passenger property loss was the most common 
complaint.  However, I did not expect to see there were four wrongful death
complaints as well as three compliments.


```{r echo=FALSE}
tsa.disposition <- tsa %>%
  group_by(Disposition) %>%
  summarize(n = n()) %>%
  arrange(desc(n))

tsa.disposition

#make waffle chart
vals <- c(96661, 47374, 33733) / 1000
val_names <- sprintf("%s (%s)", c("Deny", "Approve in Full", "Settle"), scales::percent(round(vals/sum(vals), 2)))
names(vals) <- val_names

waffle(vals,
       title = 'TSA Claims Disposition')

```

Looking at the disposition for all claims, there are 26499 missing dispositions 
(about 13% of the data). If we take out the missing data, it turns out over half
of all claims are denied, while the rest are settled or approved.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(data = tsa,
      Claim.Amount,
      binwidth = 50,
      geom = 'histogram',
      color = I('black'),
      fill = I('#CC0000')) +
  scale_x_continuous(limits = c(0, 5000), breaks = seq(0, 5000, 500))

summary(tsa$Claim.Amount)
```

Here is a histogram of claim amounts.  The data appears skewed left where 
relatively small claims seems to be more common than larger claims.  The median
claim amount is $105 while the mean is $1,003, meaning that there must be some 
very large claims that are affecting the mean.  Sure enough, we can see that the
max amount is $27,220,000.

# Univariate Analysis

### What is the structure of your dataset?

This dataset contains 204,267 different claims with 13 features for each claim:

Claim Number

Date Received

Incident Date

Airport Code

Airport Name

Airline Name

Claim Type

Claim Site

Item

Claim Amount

Status

Close Amount

Disposition

Some of these, such as Close Amount and Claim Type contain missing data.

### What is/are the main feature(s) of interest in your dataset?

The main features of interest in the dataset are date received, claim type,
claim amount, status, and close amount.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I am interested if there is any relation between an airline and claim type.

### Did you create any new variables from existing variables in the dataset?

I created three new variables by separating each date received entry by year, 
month, and day.

I also found another dataset that contained the longitude and latitude for 
airports listed by their IATA airport code.  I then joined data from that 
dataset and combined it with the data from the TSA dataset to plot the location
of each airport on the bubble map of TSA claims locations.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

There were some date received entries that had dates before 2002 and past 2015.
I fixed this by changing the date received year to be the same year as the 
incident date year.

In addition, I removed the semicolon and dollar sign from the strings for the 
claim amount and close amount and then converted it to numeric data.

I also changed all the incident dates and received dates, which were in mutiple
formats, to a single date format.

# Bivariate Plots Section

```{r echo=FALSE, warning=FALSE, message=FALSE, Bivariate_Plots}
ggplot(aes(x = date.received.year, y = Claim.Amount, group = date.received.year),
       data = subset(tsa, !is.na(Claim.Amount) & !is.na(date.received.year))) +
  ylim(0,1500) +
  scale_x_continuous(limits = c(2002, 2015),
                     breaks = seq(2002, 2015, 1)) +
  geom_boxplot()
  #scale_y_log10()
```

It appears that the median claim amount seems to decrease after 2009.  This could be related to the fact that the number of claims decreased after 2010 compared to previous years.  In October 2009, the TSA announced that it would purchase new or upgrade existing closed circuit television systems at aipports accros the country, which could be an explain the drop in claims and the drop in median claim amounts.

```{r echo=FALSE, warning=FALSE, message=FALSE}
ggplot(aes(x = Disposition, y = Claim.Amount),
       data = subset(tsa, tsa$Disposition != '', !is.na(Claim.Amount) & !is.na(date.received.year))) +
         geom_boxplot() +
  ylim(0, 75000)

ggplot(aes(x = Disposition, y = Claim.Amount),
       data = subset(tsa, tsa$Disposition != '', !is.na(Claim.Amount) & !is.na(date.received.year))) +
         geom_boxplot() +
  ylim(0, 1000)
```

These boxplots had many outliers making the plots difficult to see, which is why
I decided to take away some of the outliers by limiting the y-axis.  We can see
from the first plot that the TSA did not approve in full any claims above $1600. Furthermore, it is seems that the TSA attempted to settle or deny the most 
expensive of claims rather than approving them.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = Claim.Amount, y = Close.Amount),
       data = subset(tsa, !is.na(tsa$Claim.Amount | tsa$Close.Amount)),
       color = I('#5760AB')) +
  geom_point(fill = I('#5760AB'), color = I('black'), shape = 21, alpha = 0.25) +
  xlim(0, 100000) +
  ylim(0,100000) +
  geom_abline(intercept = 0, slope = 1, colour='#E41A1C')

cor.test(tsa$Claim.Amount, tsa$Close.Amount)

#strange.claim <- tsa[which(tsa$Claim.Amount < tsa$Close.Amount & tsa$Claim.Amount == 0),]
```

With this next plot, I was interested to see how the amount people claimed
compared to the amount they eventually received.  I included a red line to 
indicate where the amount claimed is equal to the close amount.

For very high claim amounts, very few claims recieved the full amount they asked
for.

One thing that caught my eye is that there were a few claims of $0 or near
$0 that had close amounts higher than their claim amounts.  If I had to guess
as to why this was the case, I would say that these are probably errors in the
data.

```{r echo=FALSE, message=FALSE}
tsa3 <- subset(tsa, Airline.Name != '' & tsa$Claim.Site == 'Checked Baggage') %>%
  count(Airline.Name) %>%
  top_n(8) %>%
  arrange(n, Airline.Name) %>%
  mutate(Airline.Name = factor(Airline.Name, levels = unique(Airline.Name)))

tsa %>%
  filter(Airline.Name %in% tsa3$Airline.Name) %>%
  mutate(Airline.Name = factor(Airline.Name, levels = levels(tsa3$Airline.Name))) %>%
  ggplot(aes(x = Airline.Name, fill = Claim.Type), subset(tsa3, !is.na(tsa$Claim.Type) & tsa$Claim.Type != '' & tsa$Claim.Type != '-')) +
    geom_bar() +
  coord_flip()
```

Here are the top eight airlines listed in the claims.  There doesn't seem to be 
anything interesting here, since the proportions of each claim type for each 
claim type appear pretty similar across the airlines.  Also, the top airlines 
with the most claims are also some of the airlines that fly the most amount of passengers.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = subset(tsa, !is.na(Claim.Type) & Claim.Type != '' & Claim.Type != '-'), 
  aes(x = date.received.year)) +
  geom_histogram(aes(fill = Claim.Type)) +
  scale_fill_brewer(palette="Spectral") +
  scale_x_continuous(limits = c(2002, 2015),
                     breaks = seq(2002, 2015, 1)) +
  ggtitle('Claims Types by Year')
```

Nothing too surprising with this plot as Passenger Property Loss and Property 
Damage seem to be the most common complaint across all years.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = subset(tsa, !is.na(Claim.Site) & Claim.Site != '' & Claim.Site != '-'), 
  aes(x = date.received.year)) +
  #facet_wrap(~ date.received.year) +
  geom_histogram(aes(fill = Claim.Site)) +
  scale_fill_brewer(palette="Set1") +
  scale_x_continuous(limits = c(2002, 2015),
                     breaks = seq(2002, 2015, 1)) +
  ggtitle('Claims Sites by Year')
```

The two most frequent claim sites listed for each year are checked baggage, and
checkpoint.

Counts for claims with the claim site listed as checked baggage appear to 
decrease after 2004 while counts for claims with claim site listed as checkpoint
do not change quite as much.  It is not obvious as to why this is.  It could be
that the TSA has improved its measures to prevent agents from damaging or 
stealing items from luggage, or perhaps passengers over time start to be more 
careful and pack items they do not want stolen in their carry-on luggage, or 
passengers are simply checking in less baggage because of airlines charging 
baggage fees.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = subset(tsa, !is.na(Disposition) & Disposition != '' & Disposition != '-'), 
  aes(x = date.received.year)) +
  #facet_wrap(~ date.received.year) +
  geom_histogram(aes(fill = Disposition)) +
  scale_fill_brewer(palette="Set1") +
  scale_x_continuous(limits = c(2002, 2015),
                     breaks = seq(2002, 2015, 1)) +
  ggtitle('Disposition by Year')
```

The proportion of claims denied appears to increase around 2006 or 2007.

# Bivariate Analysis

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Unfortunately, my analysis of airline and claim type did not yield any 
unexpected results.

### What was the strongest relationship you found?

The only two continuous variables I compared were claim amount and close amount.
They do not have a very strong relationship, as their correlation coefficient is
essentially zero.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
tsa$date.diff <- tsa$date.received.format - tsa$incident.date.format

qplot(Claim.Type, date.received.format - incident.date.format,
       data = subset(tsa, !is.na(date.diff) & date.diff >= 0), geom = 'boxplot') +
  coord_flip() +
  ylim(0, 200)
```

These box and whisker plots compare the number of days between the incident date
and the date the claim was received for across each type of claim.  The median 
difference in time appears fairly similar across all different claim types, 
except for possibly the compliment claim type, though since there were only four 
claims with compliments in the data, this may not be a very good indicator of
what the actual median should be.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = date.received.month, y = Close.Amount / Claim.Amount),
       data = subset(tsa, !is.na(date.diff) & Claim.Amount != 0 & !is.na(Close.Amount) & !is.na(Claim.Amount))) +
  facet_wrap(~date.received.year) +
  geom_line(stat = 'summary', fun.y = mean)
```

Here we have the average of the close amount divided by the claim amount for
each month from 2002 to 2009.  The ratio appears to be somewhat sporadic during
2002, most likely because there were relatively few claims during 2002.  

The ratio decreases over time until levelling off in 2008.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Looking at claim amount divided by close amount over time seemed to be an 
important relationship as this ratio seemed to decrease with time.

### Were there any interesting or surprising interactions between features?

It was interesting for me to see how the average claim amount over close amount
decreased over time since it shows that less money is being given out compared
to the amount being asked asked for as the years go by.
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, warning=FALSE, Plot_One}
ggmap(map) +
  geom_point(aes(x = longitude, y = latitude, color = I('#3300FF')),
             data = tsa.by.port,
             alpha = 0.25,
             size = tsa.by.port$n / 500) +
  ggtitle("Claims Locations") +
  xlab('Longitude') +
  ylab('Latitude')
```

### Description One

The airports with the most claims are tend to be heavily trafficked airports in
large cities such as JFK International Airport in New York or LAX in Los 
Angeles.

### Plot Two
```{r echo=FALSE, warning=FALSE, message=FALSE, Plot_Two}
ggplot(data = subset(tsa, !is.na(Claim.Type) & Claim.Type != '' & Claim.Type != '-'), 
  aes(x = date.received.year)) +
  geom_histogram(aes(fill = Claim.Type)) +
  scale_fill_brewer(palette="Spectral") +
  scale_x_continuous(limits = c(2002, 2015),
                     breaks = seq(2002, 2015, 1)) +
  ggtitle('Claims Types by Year') +
  xlab('Year') +
  ylab('Number of Claims')
```

### Description Two

The number of claims peak in 2004 and decrease in subsequent years.  The most 
common form of claims for all years are passenger property loss and property
damage.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = date.received.month, y = Close.Amount / Claim.Amount),
       data = subset(tsa, !is.na(date.diff) & Claim.Amount != 0 & !is.na(Close.Amount) & !is.na(Claim.Amount))) +
  facet_wrap(~date.received.year) +
  geom_line(stat = 'summary', fun.y = mean) +
  scale_x_continuous(limits = c(0, 13),
                     breaks = seq(1, 12, 1)) +
  ggtitle('Average Claim Amount/Close Amount') +
  ylab('Claim Amount/Close Amount') +
  xlab('Month')
```

### Description Three

The average ratio appears to fluctuate a lot in 2002, which is probably because
of the relatively few amount of claims during that year.  However after 2002,
changes become more gradual and over time the average claim amount/close
amount decreases and until it levels off in 2008.

------

# Reflection

### Issues

While working on this project I encountered issues using this dataset.  The most significant problem is that there were missing data.  For instance, there 
were no data for close amount after 2009, which was six years of missing data.  
Another issue was that dates were given in two different formats that needed to
be converted to a single format, which I was able to do with assistance from 
an instructor on Live Help.  Finally, data for items listed in the claims needed
to be cleaned.  Given more time, I would have liked to have explored the 
different items given and their relationships to the other variables.

One limitation of this dataset is that we do not have any data on the total 
number of travellers who pass through TSA screening.  For a more complete 
investigation, I would have liked to have seen how the number of claims compare 
to the overall count of people who are screened by the TSA over time.

Another limitation is that we cannot assess the validity of all of the claims.
It is possible that some of these claims could be fraudulent.  Furthermore, some
items stolen from passenger luggage could have been stolen by airline employees
or airport baggage handlers and the TSA was mistakenly blamed.

###Conclusion

After doing this exploratory data analysis I think air travellers will have some
mixed feelings about the results.  On one hand, I think most will be pleased to 
know that TSA claims have been decreasing since 2004, which is a sign that the 
agency may be improving its measures to prevent damage and theft of passengers' personal belongings.  On the other hand, for those who are victims of theft, the fact that over 50% of all claims between 2002 and 2015 were denied may not be a satisfying outcome.